Race Science in Mainstream Psychology, 1960-2010

### 2025-04-14 paper/paper.qmd
fix issue w/ PF author list and downstream changes in 15

In @FernandezPintoKnowBetterNot2017, the paradigm for supporting friendly research is the Tobacco Industry Research Committee, later known as the Council for Tobacco Research. @FernandezPintoKnowBetterNot2017 points out that the TIRC/CTR funded "scientific experts in major universities and research institutions" rather than marginal, fringe, or industry-employed scientists. The purpose was "to use 'good' research as a distraction" from the harmful effects of smoking by "fund[ing] research on heredity, infection, nutrition, hormones, nervous tension, and environmental factors" as causes of lung cancer and heart disease (57). This process of "industrial selection" for favorable research [@HolmanExperimentationIndustrialSelection2017] created the misleading impression that these non-tobacco causes were just as important, perhaps even more important, than smoking [compare @WeatherallHowBeatScience2018; @OConnorMisinformationAgeHow2019]. Similarly, per Laughlin, the Pioneer Fund was explicitly created to promote research on eugenics and race science as just as important, if not more important, than non-biological ("environmentalist") explanations of racial social inequality.

@@

![Disproportionate presence of Pioneer-funded authors in three focal topics.\
(A) Number of authors for three focal topics from the medium vocabulary, k= 40 model, by year. Mainstream intelligence (topic 22) shows exponential-like increase while MQ race science (topic 7) and race-and-intelligence research (topic 24) remain flat. \
(B) Pioneer-funded authors as a share (%) of all authors publishing in a given topic, by year (thin lines) and 5-year running mean (thick lines). Usually fewer than 15% of authors in a topic are Pioneer-funded. \
(C) Share of papers (%) with at least one Pioneer-funded author, by year (thin lines) and 5-year running mean (thick lines). After about 1985, more than 15% of papers in topic 24 have at least one Pioneer-funded author. \
(D) Comparison of author share (panel B) to paper share (panel C). Each point is one year-topic combination. Points above the dark line $y = x$ indicate Pioneer-funded authors had a greater paper share than would be expected based on author share. Length of thin vertical lines measures "presence." \
(E) Empirical density plots of "presence," mean difference between paper share and author share for 1,000 randomly drawn sets of 15 authors. Bold vertical lines are the mean values for the Pioneer-funded authors (signed length of lines in panel D). Pioneer-funded authors are similar to randomly drawn sets for topics 7 and 22, but very different for topic 24. ](img/15_presence.png){#fig-presence}

@@

Subfigure @fig-presence D is a scatterplot of paper share (subfigure C) against author share (subfigure B), by year and topic. The dark diagonal line $y = x$ corresponds to equality between paper share and author share. Distance above (below) this line indicates that Pioneer-funded authors had disproportionately greater (less) "presence" than would be expected based on their authorship profile. For topics 7 (MQ-style race science) and 22 (mainstream intelligence), Pioneer-funded authors consistently had a small negative to modestly positive presence. In contrast, for topic 24 (race-and-intelligence), Pioneer-funded authors consistently had a substantial, disproportionately positive presence, with in some cases differences between paper share and author share of 20 points or more.

Subfigure @fig-presence E uses a resampling method to compare the observed value of presence for the Pioneer-funded authors (dark vertical lines) to a "random" set of other authors publishing in the topic (density curves and rugs). Compared to other authors, the Pioneer-funded authors do not have a distinctive presence for topics 7 and 22, but are at the rightmost extreme for topic 24.

@@

The Crossref API does not standardize or disambiguate author names. We manually disambiguated the names of the 15 Pioneer-funded authors, e.g., searching the author field for every occurrence of "Rushton" and then identifying all the variations for J. Philippe Rushton. A simple crosswalk table allowed us to standardize all of these variations to a "canonical" version of the name of each Pioneer-funded author, e.g., J.P. Rushton would be replaced with the "canonical" version J. Philippe Rushton before proceeding with further analysis. Fig. @fig-pf-count shows the number of Pioneer-funded authors who published in the three focal topics, by year. Across the entire scope of the data, 12 of the 15 authors ever published in any of these three topics, with Rushton, Lynn, Jensen, Humphreys, and Vernon each publishing 10 or more papers (aggregated across topics).

@@

Names of other authors were not disambiguated, and so, e.g., D.J. Hicks and Daniel J. Hicks would be treated as two different authors for the purposes of this analysis. As a consequence, the total number of authors is probably too high in the estimate of `author_share`, and so the estimate of `author_share` is probably too low. Assuming no issues with the manual disambiguation of Pioneer-funded authors, `paper_share` does not have this problem. Therefore the estimated difference `paper_share - author_share` is probably larger than the true difference. In terms of subfigure #fig-presence E, the vertical bar is probably too far to the right.

@@

To contextualize this measure of "presence" within topics, we first identified all distinct authors (Pioneer-funded or otherwise, again without disambiguation for the others) across all papers in the given topic. We then used a permutation-style resampling approach: drawing a set of 15 authors and calculating paper share, author share, their difference, and the mean of this difference, exactly as with the Pioneer-funded authors. (In particular, years where author share was 0, i.e., the authors had no publications in the topic, were excluded from calculating the mean difference.)

>>>>

In @FernandezPintoKnowBetterNot2017, the paradigm for supporting friendly research is the Tobacco Industry Research Committee, later known as the Council for Tobacco Research. @FernandezPintoKnowBetterNot2017 points out that the TIRC/CTR funded "scientific experts in major universities and research institutions" rather than marginal, fringe, or industry-employed scientists (57). The purpose was "to use 'good' research as a distraction" from the harmful effects of smoking by "fund[ing] research on heredity, infection, nutrition, hormones, nervous tension, and environmental factors" as causes of lung cancer and heart disease (57). This process of "industrial selection" for favorable research [@HolmanExperimentationIndustrialSelection2017] created the misleading impression that these non-tobacco causes were just as important, perhaps even more important, than smoking [compare @WeatherallHowBeatScience2018; @OConnorMisinformationAgeHow2019]. Similarly, per Laughlin, the Pioneer Fund was explicitly created to promote research on eugenics and race science as just as important, if not more important, than non-biological ("environmentalist") explanations of racial social inequality.

@@

![Disproportionate presence of Pioneer-funded authors in three focal topics.\
(A) Number of authors for three focal topics from the medium vocabulary, k= 40 model, by year. Mainstream intelligence (topic 22) shows exponential-like increase while MQ race science (topic 7) and race-and-intelligence research (topic 24) remain flat. \
(B) Pioneer-funded authors as a share (%) of all authors publishing in a given topic, by year (thin lines) and 5-year running mean (thick lines). Usually less than 15% of authors in a topic are Pioneer-funded. \
(C) Share of papers (%) with at least one Pioneer-funded author, by year (thin lines) and 5-year running mean (thick lines). Not shown is 1963 for topic 22, where Robert A. Gordon was one of two coauthors of the single publication in that topic that year. After about 1985, more than 15% of papers in topic 24 have at least one Pioneer-funded author. \
(D) Comparison of author share (panel B) to paper share (panel C). Each point is one year-topic combination. Not shown here is 1963 for topic 22. Points above the dark line $y = x$ indicate Pioneer-funded authors had a greater paper share than would be expected based on author share. Length of thin vertical lines measures "presence." \
(E) Empirical density plots of "presence," mean difference between paper share and author share for 1,000 randomly drawn sets of 13 authors. Bold vertical lines are the mean values for the Pioneer-funded authors (signed length of lines in panel D). Pioneer-funded authors are similar to randomly drawn sets for topics 7 and 22, but very different for topic 24. ](img/15_presence.png){#fig-presence}

@@

Subfigure @fig-presence D is a scatterplot of paper share (subfigure C) against author share (subfigure B), by year and topic. The dark diagonal line $y = x$ corresponds to equality between paper share and author share. Distance above (below) this line indicates that Pioneer-funded authors had disproportionately greater (less) "presence" than would be expected based on their authorship profile. For topics 7 (MQ-style race science) and 22 (mainstream intelligence), Pioneer-funded authors consistently had a small negative to modestly positive presence; with the exception of 1963 (not shown on this plot), where Robert A. Gordon was one of two coauthors (50% author share) of the sole paper in topic 22 that year (100% author share).  In contrast, for topic 24 (race-and-intelligence), Pioneer-funded authors consistently had a substantial, disproportionately positive presence, with in some cases differences between paper share and author share of 20 points or more.

Subfigure @fig-presence E uses a resampling method to compare the observed value of presence for the Pioneer-funded authors (dark vertical lines) to a "random" set of 13 other authors publishing in the topic (density curves and rugs). Compared to other authors, the Pioneer-funded authors do not have a distinctive presence for topic 7, are somewhat greater than the mode for topic 22, and are at the rightmost extreme for topic 24.

@@

The Crossref API does not standardize or disambiguate author names. We manually disambiguated the names of the Pioneer-funded authors, e.g., searching the author field for every occurrence of "Rushton" and then identifying all the variations for J. Philippe Rushton. A simple crosswalk table allowed us to standardize all of these variations to a "canonical" version of the name of each Pioneer-funded author, e.g., J.P. Rushton would be replaced with the "canonical" version J. Philippe Rushton before proceeding with further analysis. Fig. @fig-pf-count shows the number of Pioneer-funded authors who published in the three focal topics, by year. Across the entire scope of the data, 13 of the Pioneer-funded authors ever published in any of these three topics, with Rushton, Lynn, Jensen, Humphreys, and Vernon each publishing 10 or more papers (aggregated across topics).  (Note that Lloyd Humphreys and Robert A. Gordon did not have any results in the Web of Science searches but did have papers in the corpus. Hardin, Horn, and Shuey do not appear in the papers for these focal topics.)

@@

Names of other authors were not disambiguated, and so, e.g., D.J. Hicks and Daniel J. Hicks would be treated as two different authors for the purposes of this analysis. As a consequence, the total number of authors is probably too high in the estimate of `author_share`, and so the estimate of `author_share` is probably too low. Assuming no issues with the manual disambiguation of Pioneer-funded authors, `paper_share` does not have this problem. Therefore the estimated difference `paper_share - author_share` is probably larger than the true difference. In terms of subfigure @fig-presence E, the vertical bar is probably too far to the right.

@@

To contextualize this measure of "presence" within topics, we first identified all distinct authors (Pioneer-funded or otherwise, again without disambiguation for the others) across all papers in the given topic. We then used a permutation-style resampling approach: drawing a set of 13 authors and calculating paper share, author share, their difference, and the mean of this difference, exactly as with the Pioneer-funded authors. (In particular, years where author share was 0, i.e., the authors had no publications in the topic, were excluded from calculating the mean difference.)
